Supplementary Information

### 2020-11-25 doc/manuscript/MEE_manuscript.Rmd
worked on ms

```{r eval = TRUE, echo=FALSE, warning=FALSE, results='hide'}
#Tables
sites = 1
#Figures
schematic=1
SCBI_example=2
clim_sens=3
size=4
full_examples = 5
# Fig5/ table2- DBH/climate interactions

@@

**Here, we develop a new method that allows simultaneous consideration of the effects of tree size and objectively determined principle climate drivers on annual tree growth.** 
This allows us to ask:
(1) What are the most important climate drivers of annual growth, and over which time windows?
(2) What is the shape of the relationship between annual growth and climate drivers?
(3) How do $RW$, $BAI$, and aboveground biomass increments ($\Delta AGB$) vary with $DBH$?
(4) How do climate drivers and $DBH$ additively and interactively shape growth?

@@

knitr::kable(sites[, c("site code",	"site name",	"location",	"July T",	"Jan T",	"Ann. Precip.",	"vegetation type(s)",	"n species",	"n cores",	"original publication(s)")], 
                  col.names = c("site code",	"site name",	"location",	"July T",	"Jan T",	"Ann. Precip.",	"vegetation type(s)",	"n species",	"n cores",	"original publication(s)"), 
                  caption = "Table 1. Sites included in this analysis.") 
```

@@

**We used the *climwin* package in R [@van_de_pol_identifying_2016] to identify the most important climate driver and the time window over which its effect was strongest for each of two categories of variables: a temperature group (mean, min, and max temperature; PET) and a precipitation group (precipitation, number of days with precipitation).**
To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by using a generalized additive model (GAM) to fit a spline to individual growth records ($RW$, $BAI$, or $\Delta AGB$) from each core, thereby producing residuals. (*How did you decide how flexible to make the spline?*)
We then used *climwin* to identify the climate variables most strongly correlated to the residuals of $RW$, $BAI$, or $\Delta AGB$, specifying quadratic fits to allow for potential nonlinearities in the climate response.
Within *climwin*, we specified a mixed effects model using species and core identity as random effects: `residual growth index ~ [climate] + (1 | sp) + (1 | treeID)`.
We allowed first-order linear or concave-down second-order polynomial fits, but ignored any concave-up fits on the basis that these are not expected biologically and, when identified in initial analyses, often appeared to be cases of over-fitting (**[ISSUE #84 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/84)**).

Here, for each permutation, `climate` specifies one of the climate drivers in the climate variable group, analyzed over one of all possible combinations of consecutive months over a 15 month period ending near the time of cessation of formation of each annual ring (Table `r site_details`). (**[ISSUE #51 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/51)**)
*Climwin* runs all potential models to select the best fit (lowest AIC), and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@van_de_pol_identifying_2016]. 
For each group of candidate climate variables, we move forward with the best variable over the time window identified by *climwin* as a candidate climate variable for the multivariate models.

A challenge to this system arose for the sites that have undergone the most rapid changes in climate and tree growth: SC and LT, where trees exhibit significant growth declines attributed to rising temperatures [@sniderhan_growth_2016] and increasing drought (REF), respectively. 
*([ISSUE #25 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25))*
Problematically, correlating tree growth residuals from which climate-driven trends had been removed against the climate signal with a strong directional trend would not necessarily identify the most relevant climate drivers.
For these sites, we experimented with three approaches to identifying the most important climate drivers (1) the method described above, (2) detrending the climate variables **(AT:prewhitening?)** prior to the climwin step, and (3) excluding *decades with the most pronounced climate change* from the climwin step of the analysis [*sensu*  @sniderhan_growth_2016] (Appendix `r changing_climate`).
After determining that... , here we present results *(method)*.

We verified that this process identified similar climate variable-month combinations as what would be identified using traditional methods for individual species, as detailed in Appendix `r traditional_comparison`. (**[ISSUE #35 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/35)**)

@@

Having identified candidate climate drivers in temperature and precipitation variable groups, we next combined climate variables (all models) and $DBH$ (models with $DBH$ and its climate interactions) in a generalized least squares (GLS) model (Fig. `r schematic`). 
Before running the models, we checked for collinearity among the candidate variables using the *vifstep* function (**REF**) and removed any variable with a variance inflation factor > 3 (none required removal).
Within the GLS models, our response variables were $log[\Delta r]$, $log[BAI]$, or $log[\Delta AGB]$. 
Rather than detrending these variables to produce residuals, the temporal autocorrelation of individual tree's growth was accounted for by the specifying an autocorrelation structure of order 1, with $Year$ as a continuous time covariate and $coreID$ as a grouping factor, in the GLS's model specification.
For each species independently, we ran every combination of the candidate climate variables and $DBH$, including both first- and second-order terms of polynomial for each.
For models including interactive effects of climate and $DBH$, we included only first-order linear terms for both $DBH$ and climate variables. (*[ISSUE #42 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/42)*)
Within each of three categories of models run (climate only, $DBH$+climate, $DBH$ x climate), we selected as the top model that with the lowest AIC.

@@

**Our process picked out similar climate drivers to what would be obtained via traditional methods (Figs. `r SCBI_example`, `r all_traditional_comparisons`; Table `r qualitative_results_comparison`; Appendix `r traditional_comparison`), but with the advantage that *climwin* allowed automated, objective selection of the strongest climate drivers and the time windows over which they were most influential.**
The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, each of which was identified by climwin as the top temperature-related driver *at four of the eight sites*. $T_{min}$ was the top driver at only *one site (BCI)*, and $T_{mean}$ was never identified as the top variable within the temperature group (Fig. `r clim_sens`).
Within the precipitation group, precipitation amount ($PPT$) was identified as the top variable most frequently (n=*6 of 9* sites), but it was not uncommon that it was surpassed by precipitation frequency ($PDF$; n=*3 of 8* sites). 
Optimal time windows often coincided with a site's peak growing season (n= *# of 10* for temperature variables, *# of 10* for precipitation variables), but exceptions were common. 
At the *5 lowest latitude* sites (BCI, HKK, LT, CB, and SCBI), the optimal window for precipitation variables spanned $\ge$ 8 months, ending during the peak growing months of the year of ring formation. 
At the *3 highest latitude sites* (HF, ZOF, and SC), the optimal window for precipitation variables was a short ($\le$ 3 months) window during the previous growing season.
Optimal windows for temperature variables tended to be shorter, the longest being a *6*-month period during the summer (wet season) at HKK. 
At two of the higher-latitude temperate sites (HF and Žofín), temperatures were most influential during late winter/ early spring.
There were also a few instances where previous growing season conditions had the strongest influence.

@@

@@

Precipitation responses were included in the best model at all sites and *for all but # species* (Fig. `r clim_sens`).
Responses were most commonly positive, and were most pronounced at the driest site (LT). 
Precipitation terms included in top models were non-linear #% of the time, and significantly better than first-order linear model #% of the time.
In some cases, the non-linearity was quite pronounced (e.g., LT), with the most common pattern (#%) being a decelerating increase.

@@

# Discussion

@@

## corroboration of insight from conventional methods
*( more objective approach to produce similar results to what has been shown using more conventional methods)*
*(for identigying climate variables... Is the benefit of the new approach that it is a more objective way to reach the same interpretation, or are there cases when it would produce a different interpretation?)*

@@

@@

Sets the foundation for considering other, slowly changing environmental drivers.

>>>>

```{r eval = TRUE, echo=FALSE, warning=FALSE, results='hide'}
#Tables
sites = 1
#Figures
schematic=1
SCBI_example=2
clim_sens=3
size=4
full_examples = 5
# Fig5/ table2- DBH/climate interactions
year_fig = 7

@@

**Here, we develop a new method that allows simultaneous consideration of the effects of tree size and objectively determined principle climate drivers on annual tree growth.** 
This allows us to ask:
(1) What are the most important climate drivers of annual growth, and over which time windows?
(2) What is the shape of the relationship between annual growth and climate drivers?
(3) How do $RW$, $BAI$, and aboveground biomass increments ($\Delta AGB$) vary with $DBH$?
(4) How do climate drivers and $DBH$ additively and interactively shape growth? 
(5) How have growth rates changed through time?

@@

knitr::kable(sites[, c("site code",	"site name",	"location",	"July T",	"Jan T",	"Ann. Precip.",	"vegetation type(s)",	"n species",	"n cores",	"original publication(s)")], 
                  col.names = c("site code",	"site name",	"location",	"July T",	"Jan T",	"Ann. PPT",	"vegetation type(s)",	"n species",	"n cores",	"original publication(s)"), 
                  caption = "Table 1. Sites included in this analysis.") 
```

@@

**We used the *climwin* package in R [@van_de_pol_identifying_2016] to identify the most important climate driver and the time window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and $PET$) and a precipitation group ($PPT$, $PDF$).**
To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by using a generalized additive model (GAM) to fit a spline to individual growth records ($RW$, $BAI$, or $\Delta AGB$) from each core, thereby producing residuals. (*How did you decide how flexible to make the spline?*)
We then used *climwin* to identify the climate variables most strongly correlated to the residuals of $RW$, $BAI$, or $\Delta AGB$, specifying quadratic fits to allow for potential nonlinearities in the climate response.
Within *climwin*, we specified a mixed effects model using species and core identity as random effects: `residual growth index ~ [climate] + (1 | sp) + (1 | treeID)`.
Here, for each permutation, `climate` specifies one of the climate drivers in the climate variable group, analyzed over one of all possible combinations of consecutive months over a 15 month period ending near the time of cessation of formation of each annual ring (Table `r site_details`). 
*Climwin* runs all potential models to select the best fit (lowest AIC), and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@van_de_pol_identifying_2016]. 
For each group of candidate climate variables, we moved forward with the best variable over the time window identified by *climwin* as a candidate climate variable for the multivariate models.

We tested whether this process identified similar climate variable-month combinations as what would be identified using traditional methods for individual species, as detailed in Appendix `r traditional_comparison`. 
Furthermore, we explored alternate methods of climate variable selection for the two sites that have undergone the most rapid changes in climate and tree growth: SC, where rapidly rising temperatures are causing melting permafrost, summer moisture stress, and growth declines [@sniderhan_growth_2016], and LT, where increasingly warm drought has dramatically reduced growth [@williams_temperature_2012]. 
We ultimately determined that the method described above was reliable for the purposes of this analysis (Appendix `r changing_climate_title`).

@@

Having identified candidate climate drivers in temperature and precipitation variable groups, we next combined climate variables (all models) and $DBH$ (models with $DBH$ and its climate interactions) in a generalized least squares (GLS) model (Fig. `r schematic`). 
Before running the models, we checked for collinearity among the candidate variables using the *vifstep* function (**REF**) and removed any variable with a variance inflation factor > 3 (none required removal).
Within the GLS models, our response variables were $log[\Delta r]$, $log[BAI]$, or $log[\Delta AGB]$. 
Rather than detrending these variables to produce residuals, the temporal autocorrelation of individual tree's growth was accounted for by the specifying an autocorrelation structure of order 1, with $Year$ as a continuous time covariate and $coreID$ as a grouping factor, in the GLS's model specification.
For each species independently, we ran every combination of the candidate climate variables and $DBH$, including both first- and second-order terms of polynomial for each.
For climate responses, we allowed concave-down fits, but ignored any concave-up fits on the basis that these are not expected biologically and, when identified in initial analyses, often appeared to be cases of over-fitting.
For models including interactive effects of climate and $DBH$, we included only first-order linear terms for both $DBH$ and climate variables.
Within each of three categories of models run (climate only, $DBH$+climate, $DBH$ x climate), we selected as the top model that with the lowest AIC.

@@

**Our process picked out similar climate drivers to what would be obtained via traditional methods (Figs. `r SCBI_example`, `r all_traditional_comparisons`; Table `r qualitative_results_comparison`; Appendix `r traditional_comparison`), with the advantage that *climwin* allowed automated selection of the strongest climate drivers and the time windows over which they were most influential.**
Within the precipitation group, precipitation amount ($PPT$) was identified as the top variable most frequently (n=*7 of 10* sites), but was sometimes surpassed by precipitation frequency ($PDF$; n=*3 of 10* sites). 
The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, which were identified by *climwin* as the top temperature-related driver *at five and four of the ten sites, respectively*. $T_{min}$ was the top driver at only *one site (BCI)*, where it's effects were only marginally significant, and $T_{mean}$ was never identified as the top variable within the temperature group (Fig. `r clim_sens`).

@@

Optimal time windows often coincided with a site's peak growing season (n= *# of 10* for temperature variables, *# of 10* for precipitation variables), but exceptions were common. 
At the *5 lowest latitude* sites (BCI, HKK, LT, CB, and SCBI), the optimal window for precipitation variables spanned $\ge$ 8 months, ending during the peak growing months of the year of ring formation. 
At the *3 highest latitude sites* (HF, ZOF, and SC), the optimal window for precipitation variables was a short ($\le$ 3 months) window during the previous growing season.
Optimal windows for temperature variables tended to be shorter, the longest being a *6*-month period during the summer (wet season) at HKK. 
At two of the higher-latitude temperate sites (HF and Žofín), temperatures were most influential during late winter/ early spring.
There were also a few instances where previous growing season conditions had the strongest influence.

@@

Precipitation responses ($PPT$ or $PDF$) were included in the best model at all sites and *for all but # species* (Fig. `r clim_sens`).
Responses were most commonly positive, and were most pronounced at the driest site (LT). 
Increases in growth rate decelerated or decreased at higher $PPT$ or $PDF$ (*i.e.*, the best model contained a negative second-order term) for #% of site-species combinations.

@@

### Effects of year
Growth rates changed directionally over time--i.e., there was a significant effect of `year` in the GLS model-- for # species at # sites (Table **XX**).
In the majority of cases (#%), the trend was negative.
Declines were particularly dramatic (**give some criteria**) for *all four species at Harvard Forest, PIPO at Little Tesuque, POTR at Cedar Breaks, and PIMA at Scotty Creek* (Fig. `r year_fig`).
In these cases, models without `year` predicted declines in growth at large DBH that should more propoerly be attributed to the effect of year (Fig. `r year_fig`).

![**Figure 7 | DRAFT FIGURE. GLS model with and without year, maybe decadal analysis plot. IT WOULD BE BEST TO COMBINE THE 2 DBH FIGURES (PANELS 1,2). CURRENT EXAMPLE IS JUST SCOTTY CREEK, BUT IT WOULD BE GOOD TO SHOW AT LEAST ONE OTHER.](tables_figures/year_figure_sketch.png)
# Discussion

@@

## corroboration of insight from conventional methods
*( detecting similar climate sensitivity to  conventional methods-- identical not expected)*

  *(for identigying climate variables... Is the benefit of the new approach that it is a more objective way to reach the same interpretation, or are there cases when it would produce a different interpretation?)*

*(size patterns consistent with "textbook" for sun trees, consistent with known ecology)*

@@

- non-linearities exist
- interactions exist
- growth rates changing directionally?

@@

-  Sets the foundation for considering other, slowly changing environmental drivers.

- Framework can be used with other variables.

# Misc content/ notes to integrate
- Scotty temperture responses are probably not physiological, but due to permafrost.
